The instability of variational quantum mechanics with Gaussian approximants is (not) entirely due to the unstable least squares problem

### 2013-06-11 paper.tex
task

>>>>

The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian $$H=\hbar a^{2\dagger}a²$.  It will do so by the midpoint method, discretising the differential equation
$$Du(t)=f(t)$$
to the difference equation
$$v^{n+1}=v^{n-1}+2τfⁿ,$$
where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  The results will be compared for three difference choices of $v$ and $f$.

In the first choice, the quantum state at time $t$ is expanded over a truncated Fock basis $|N〉$, as 
$$|ν(c(t))〉=|N〉c(t).$$
Schrödinger's equation expands to
$$|N〉Dc(t)=|D(ν\circ c)(t)〉≈{H\over i\hbar}|ν(c(t))〉≈|N〉〈N|{H\over i\hbar}|N〉c(t),$$
whence
$$Dc(t)≈〈N|{H\over i\hbar}|N〉c(t).$$
In this method, we set
$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=〈N|{H\over i\hbar}|N〉c(t),$$
solving Schrödinger's equation quite conventionally.

In the second choice, the state is approximated as a superposition of coherent states,
$$|ψ(z(t))〉=∑_{φ,α∈z(t)}e^{φ+αa†}|0〉=∑_{φ,α∈z(t)}e^{φ+½|α|²}|α〉.$$
Time is discretised immediately, setting
$$vⁿ=|ψ(zⁿ)〉\qquad{\rm and}\qquad fⁿ={H\over i\hbar}|ψ(zⁿ)〉,$$
so that the midpoint formula becomes
$$|ψ(z^{n+1})〉≈|ψ(z^{n-1})〉+{2τH\over i\hbar}|ψ(z^ⁿ)〉.$$
The approximation is due to basis truncation, as always in a least squares sense.  Writing the unknown $|ψ(z^{n+1})〉$ in terms of $|Dψ(zⁿ)〉$ and expanding over a Fock basis gives
$$ 〈N|Dψ(zⁿ)〉\bigl(z^{n+1}-zⁿ\bigr)≈〈N|ψ(z^{n-1})〉-〈N|ψ(zⁿ)〉-2iτ〈N|H|ψ(zⁿ)〉.$$
This matrix equation can be solved for $z^{n+1}-zⁿ$, which can be used to update $z$.

The third choice, the conventional variational one, is to discretise the state first, setting
$$vⁿ=zⁿ\qquad{\rm and}\qquad |Dψ(zⁿ)〉fⁿ≈{H\over i\hbar}|ψ(zⁿ)〉.$$
The midpoint formula is used exactly, and $fⁿ$ satisfies the least squares problem
$$〈N|Dψ(zⁿ)〉fⁿ≈-i〈N|Dψ(zⁿ)〉{H\over \hbar}|ψ(zⁿ)〉.$$
Usually, this would be expanded over $〈Dψ(zⁿ)|$ to give a set of normal equations.  However, when $|Dψ(zⁿ)〉$ is near singular, the numerical stability of the normal equations is suspect.  Expanding over an orthonormal basis removes this confounding instability.

To do this, we need some bracket matrices.  We have
$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
The bracket with the number states is easily evaluated:
$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
$$〈n|H|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
